---
title: "Quality Control: {{name}}"
author: "`r getOption('author')`"
date: "`r Sys.Date()`"
params:
  design: !r formula("{{design}}")
  data_dir: !r file.path("../data", Sys.Date())
  results_dir: !r file.path("../results", Sys.Date(), "quality_control")
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE}
bcbioRNASeq::prepareRNASeqTemplate()
library(tximport)
library(QoRTs)
source("_setup.R")


# install.packages("http://hartleys.github.io/QoRTs/QoRTs_LATEST.tar.gz",
#                   repos=NULL,
#                  type="source");
# Directory paths ==============================================================
invisible(mapply(
  FUN = dir.create,
  path = c(params$data_dir, params$results_dir),
  MoreArgs = list(showWarnings = FALSE, recursive = TRUE)
))
```

```{r samples}
samples <- read.csv("../sample_info/{{sample_sheet}}")
```

```{r annotations}
# Read in gencode annotations
gtf <-
  rtracklayer::import(
    "{{gtf}}"
  )

names(gtf@elementMetadata@listData)[which(names(gtf@elementMetadata@listData) == "gene_id")] <- "geneID"
names(gtf@elementMetadata@listData)[which(names(gtf@elementMetadata@listData) == "gene")] <-
  "geneName"
names(gtf@elementMetadata@listData)[which(names(gtf@elementMetadata@listData) == "gene_biotype")] <-
  "geneBiotype"
gtf@elementMetadata$id <- gtf@elementMetadata$geneID

gtfdf <- gtf %>%
  as.data.frame()

g2s <- gtf@elementMetadata@listData %>%
  as.data.frame() %>%
  select(geneID, geneName) %>%
  distinct()

tx2gene <- gtfdf %>%
  mutate(TXNAME = transcript_id, GENEID = geneID) %>%
  select(TXNAME, GENEID)
```


# Count matrices

```{r tximport}
# Read in the salmon counts data.
files <- file.path("../results", "salmon", samples$SampleID, "quant.sf")
names(files) <- samples$SampleID
txi.salmon <- tximport(files, type = "salmon", tx2gene = tx2gene)
```


```{r deseq}
# Make sure the columns are in the proper order.
rownames(samples) <- samples$SampleID
samples <- samples[colnames(txi.salmon$counts), ]
stopifnot(all(rownames(samples) == colnames(txi.salmon$counts)))

# Make sure the rows are in the proper order.
names(gtf) <- gtf$geneID
gtf <- gtf[rownames(txi.salmon$counts)]

# Create DESeq2 object
dds <- DESeq2::DESeqDataSetFromTximport(txi.salmon,
  colData = samples,
  design = params$design,
  rowRanges = gtf
)
```

```{r counts}
# Raw counts (don't use for plotting)
raw_counts <- txi.salmon$counts

# DESeq2 normalized counts
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized = TRUE)

# Transcripts per million
tpm <- txi.salmon$abundance

saveData(raw_counts, normalized_counts, tpm, dir = params$data_dir)
writeCounts(raw_counts, normalized_counts, tpm, dir = params$results_dir)
```


- [`normalized_counts.csv.gz`](`r file.path(params$results_dir, "normalized_counts.csv.gz")`): Use to evaluate individual genes and/or generate plots. These counts are normalized for the variation in sequencing depth across samples.
- [`tpm.csv.gz`](`r file.path(params$results_dir, "tpm.csv.gz")`): Transcripts per million, scaled by length and also suitable for plotting.
- [`raw_counts.csv.gz`](`r file.path(params$results_dir, "raw_counts.csv.gz")`): Only use to perform a new differential expression analysis. These counts will vary across samples due to differences in sequencing depth, and have not been normalized. Do not use this file for plotting genes.

```{r}
# Parse multiqc
multiqc <- readJSON("{{multiqc_data}}")
multiqc_general_stats <- multiqc$report_general_stats_data

read_multiqc_general_stats <- function(multiqc_stats, index, samples) {
  gs <- multiqc_stats[[index]]

  gstat <- as.data.frame(t(unlist(gs[[1]])))
  gstat$SampleID <- names(gs)[[1]]
  for (i in 2:length(gs)) {
    tmp <- as.data.frame(t(unlist(gs[[i]])))
    tmp$SampleID <- names(gs)[[i]]
    gstat <- bind_rows(gstat, tmp)
  }
  gstat <- merge(gstat, samples)
  return(gstat)
}

stats <- c()
for (j in 1:length(multiqc$report_general_stats_data)) {
  stats <- c(
    stats,
    list(read_multiqc_general_stats(
      multiqc_general_stats,
      j,
      samples
    ))
  )
}

# TODO: These might not always be the same modules or in the same order.
# Need to get these from somewhere else.
names(stats) <- c("featureCounts", "salmon", "STAR")

interestingGroups <- "FeedingState"
```


# Read metrics {.tabset}

## Total reads

High quality RNA-seq samples ideally should have at least 10 million reads per sample.

```{r plot_total_reads}
g <- ggplot(stats[["STAR"]], aes(
  x = SampleID,
  y = total_reads / 1e6L,
  fill = !!sym(interestingGroups)
))
g + geom_bar(stat = "identity") +
  labs(
    title = "total reads",
    x = NULL,
    y = "reads per million",
    fill = paste(interestingGroups, collapse = ":\n")
  ) +
  bcbio_geom_abline(yintercept = 10) +
  coord_flip() +
  scale_fill_manual(values = colorspace::lighten(pals::watlington(4),
    amount = 0.2
  ))
```

## Mapped reads

The number of mapped reads should correspond to the number of total reads.

```{r plot_mapped_reads}
# TODO: Should plot uniquely mapped only or all mapped reads. Also,
# are all multimapped reads kept? What is the STAR setting for this
# and how does salmon deal with it?
g <- ggplot(stats[["STAR"]], aes(
  x = SampleID,
  y = uniquely_mapped / 1e6L,
  fill = !!sym(interestingGroups)
))
g + geom_bar(stat = "identity") +
  labs(
    title = "uniquely mapped reads",
    x = NULL,
    y = "reads per million",
    fill = paste(interestingGroups, collapse = ":\n")
  ) +
  bcbio_geom_abline(yintercept = 10) +
  coord_flip() +
  scale_fill_manual(values = colorspace::lighten(pals::watlington(4),
    amount = 0.2
  ))
```

## Mapping rate

The genomic mapping rate represents the percentage of reads mapping to the reference genome. Low mapping rates are indicative of sample contamination, poor sequencing quality or other artifacts.

```{r plot_mapping_rate}
p <- ggplot(stats[["STAR"]], aes(
  x = SampleID,
  y = uniquely_mapped_percent + multimapped_percent,
  fill = !!sym(interestingGroups)
))
p + geom_bar(stat = "identity") +
  labs(
    title = "mapping rate",
    x = NULL,
    y = "mapping rate (%)",
    fill = paste(interestingGroups, collapse = ":\n")
  ) +
  bcbio_geom_abline(yintercept = 90) +
  coord_flip() +
  scale_fill_manual(values = colorspace::lighten(pals::watlington(4),
    amount = 0.2
  )) +
  ylim(0, 100)
```

## Number of genes detected

```{r plot_genes_detected}
min_counts <- 1L

geneCount <- colSums(raw_counts >= min_counts)
stats[["STAR"]]$geneCount <- geneCount

p <- ggplot(stats[["STAR"]], aes(
  x = SampleID,
  y = geneCount,
  fill = !!sym(interestingGroups)
))
p + geom_bar(stat = "identity") +
  labs(
    title = "genes detected",
    x = NULL,
    y = "gene count",
    fill = paste(interestingGroups, collapse = ":\n")
  ) +
  coord_flip() +
  scale_fill_manual(values = colorspace::lighten(pals::watlington(4),
    amount = 0.2
  ))
```


```{r}
library(dupRadar)
dm <- analyzeDuprates(bam = "{{duplicate_marked_bam}}",
                      gtf = "{{gtf}}",
                      paired = TRUE,
                      threads = 4)

par(mfrow=c(1,2))
duprateExpDensPlot(DupMat=dm)
duprateExpBoxplot(DupMat=dm)
duprateExpPlot(DupMat=dm,addLegend=FALSE)
duprateExpDensPlot(dm.gc[dm.gc$gc<=45,], main="below median GC genes")
```

## Gene detection saturation

We should observe a linear trend in the number of genes detected with the number of mapped reads, which indicates that the sample input was not overloaded.

```{r plot_gene_saturation}
p <- ggplot(stats[["STAR"]], aes(
  x = uniquely_mapped / 1e6L,
  y = geneCount,
  color = !!sym(interestingGroups)
))
p + geom_point() +
  labs(
    title = "genes detected",
    x = "mapped reads per million",
    y = "gene count",
    fill = paste(interestingGroups, collapse = ":\n")
  ) +
  scale_color_manual(values = colorspace::lighten(pals::watlington(4),
    amount = 0.2
  ))
```

## Exonic mapping rate

Ideally, at least 60% of total reads should map to exons.

```{r plot_exonic_mapping_rate}
# TODO: I don't know where to get this data. Actually it seems that it's from
# qualimap
# plotExonicMappingRate(bcb)
```

## Intronic mapping rate

The majority of reads should map to exons and not introns.

```{r plot_intronic_mapping_rate}
# TODO: I don't know where to get this data.
# plotIntronicMappingRate(bcb)
```

```{r plot_rrna_mapping_rate, eval="rrnaRate" %in% colnames(metrics(bcb)), echo=FALSE}
mdHeader("rRNA mapping rate", level = 2)
asis_output("Samples should have a ribosomal RNA (rRNA) contamination rate below 10%.")

# TODO: I'm unsure if this is the best way to do this.
# Certainly it is the easiest way, especially for python,
# since I don't know how to track down the rRNA fasta for
# python.
# It seems that I can use qualimap in the future to achieve this stat.
rrna_biotypes <- c("rRNA", "Mt_rRNA", "tRNA", "MT_tRNA")

rrna_gtf <- gtf[which(gtf$geneBiotype %in% rrna_biotypes)]
stopifnot(rrna_gtf$type == "gene")
rrna_counts <- raw_counts[which(rownames(raw_counts) %in% rrna_gtf$geneID), ]
stats[["STAR"]]$rrna_rate <- colSums(rrna_counts) / colSums(raw_counts)

p <- ggplot(stats[["STAR"]], aes(
  x = SampleID,
  y = rrna_rate * 100,
  fill = !!sym(interestingGroups)
))
p + geom_bar(stat = "identity") +
  labs(
    title = "rRNA mapping rate",
    x = NULL,
    y = "rRNA mapping rate (%)",
    fill = paste(interestingGroups, collapse = ":\n")
  ) +
  bcbio_geom_abline(yintercept = 10) +
  coord_flip() +
  scale_fill_manual(values = colorspace::lighten(pals::watlington(4),
    amount = 0.2
  )) +
  ylim(0, 100)

# plotRRNAMappingRate(bcb)
```

## 5'->3' bias

RNA-seq data can have specific biases at either the 5’ or 3’ end of sequenced fragments. It is common to see a small amount of bias, especially if polyA enrichment was performed, or if there is any sample degradation. If a large amount of bias is observed here, be sure to analyze the samples with a Bioanalyzer and check the RIN scores.

```{r plot_5prime_3prime_bias}
plot5Prime3PrimeBias(bcb)
stats[[3]]
# PF is Illumina terminology.
# In SAM-speak, it corresponds to the absence of flag 0x200
```

## Counts per gene

Generally, we expect similar count spreads for all genes between samples unless the library sizes or total RNA expression are different.

We recommend visualizing counts normalized with the Trimmed Mean of M-Values (TMM) method here [@Robinson2010-np]. TMM normalization equates the overall expression levels of genes between samples under the assumption that the majority of them are not differentially expressed. Therefore, by normalizing for total RNA expression by sample, we expect the spread of the TMM-normalized counts per gene to be similar for every sample.

```{r plot_counts_per_gene}
plotCountsPerGene(bcb, normalized = "tmm")
```

```{r plot_biotype, eval="geneBiotype" %in% colnames(rowData(bcb)), echo=FALSE}
mdHeader("TPM per biotype", level = 2)
asis_output("Different RNA-seq processing methods can preferentially capture a subset of the RNA species from the total RNA. For example, polyA selection should select for mostly coding genes and skip a large percentage of non-polyA non-coding RNA. Here we make boxplots of the TPM for the top 12 biotypes with the most genes assigned to them for each sample.")

keep_biotypes <- rowData(bcb) %>%
  as.data.frame() %>%
  group_by(geneBiotype) %>%
  summarise(nBiotype = n()) %>%
  arrange(-nBiotype) %>%
  top_n(12, wt = nBiotype) %>%
  pull(geneBiotype) %>%
  droplevels()
biotype_tpm <- tpm(bcb) %>%
  as.data.frame() %>%
  rownames_to_column("geneID") %>%
  gather(key = sampleID, value = tpm, -geneID) %>%
  left_join(as.data.frame(rowData(bcb)), by = "geneID") %>%
  filter(geneBiotype %in% keep_biotypes) %>%
  filter(tpm > 0)

ggplot(
  data = biotype_tpm,
  mapping = aes(
    x = sampleID,
    y = tpm,
    fill = sampleID
  )
) +
  geom_violin(
    color = "black",
    scale = "area"
  ) +
  scale_y_log10() +
  facet_wrap(~geneBiotype, scales = "free_y") +
  labs(
    title = "tpm per biotype",
    x = NULL,
    y = "transcripts per million (tpm)"
  ) +
  guides(fill = FALSE) +
  theme(axis.text.x = element_text(angle = 90L, hjust = 1L, vjust = 0.5))
```

```{r plot_broad_class, eval="broadClass" %in% colnames(rowData(bcb)), echo=FALSE}
mdHeader("TPM per broad biotype class", level = 2)
asis_output("The Ensembl biotype clasifications are too specific to plot them all. Here we have grouped the biotypes into broad classes and plot boxplots of the TPM for each sample.")

broad_tpm <- tpm(bcb) %>%
  as.data.frame() %>%
  rownames_to_column("geneID") %>%
  gather(key = sampleID, value = tpm, -geneID) %>%
  left_join(as.data.frame(rowData(bcb)), by = "geneID") %>%
  filter(!is.na(broadClass)) %>%
  filter(tpm > 0)

ggplot(
  data = broad_tpm,
  mapping = aes(
    x = sampleID,
    y = tpm,
    fill = sampleID
  )
) +
  geom_violin(
    color = "black",
    scale = "area"
  ) +
  facet_wrap(~broadClass, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "tpm per broad biotype class",
    x = NULL,
    y = "transcripts per million (tpm)"
  ) +
  guides(fill = FALSE) +
  theme(axis.text.x = element_text(angle = 90L, hjust = 1L, vjust = 0.5))
```

## Count density

Generally, we expect similar count spreads for all genes between samples unless the total expressed RNA per sample is different.

```{r plot_count_density}
plotCountDensity(
  object = bcb,
  interestingGroups = "sampleName",
  normalized = "tmm"
)
```

# Fit modeling

Several quality metrics are first assessed to explore the fit of the model, before differential expression analysis is performed.

```{r identical_samples?, echo=FALSE, results='hide', message=FALSE}
## If there are no replicates, or samples are identical, don't plot dispersions
dds <- as(bcb, "DESeqDataSet")
identical_samples <- all(rowSums(assay(dds) == assay(dds)[, 1]) == ncol(dds))
```

```{r plot_disp_ests, eval=!identical_samples}
mdHeader("Dispersion", level = 2)
asis_output("The following plot shows the dispersion by mean of normalized counts. We expect the dispersion to decrease as the mean of normalized counts increases.")
plotDispEsts(bcb)
```

## Variance stabilization

These plots show the standard deviation of normalized counts using `log2()`, `varianceStabilizingTransform()`, `rlog()`, and `tmm()` by `rank(mean)`. Note that all counts shown are log2 scale.

```{r plot_mean_sd}
plotMeanSD(bcb)
```

# Sample similarity analysis

Before performing similarity analysis, we transform counts to log2, which acts to minimize large differences in sequencing depth and helps normalize all samples to a similar dynamic range. For RNA-seq count data, variance increases with the mean. Logarithmic transformation of normalized count values with a small pseudocount will account for large variations seen between the highest expressing genes so that these genes won't dominate the PCA plots. However, due to the strong noise among low count values due to Poisson, the general log2 transformation will amplify this noise, and instead, low count genes will now dominate the PCA plots. So instead, we use a regularized log ("rlog"; base 2) transformation that gives similar results for high counts as a log2 transformation but also shrinks the values of low counts towards the genes’ average across samples. We do this with the `rlog()` function in the [DESeq2][] package [@Love2014-sq], which we will later use for differential gene expression analysis.

## Principal component analysis (PCA)
PCA [@Jolliffe2002-jz] is a multivariate technique that allows us to summarize the systematic patterns of variations in the data. PCA takes the expression levels for genes and transforms it in principal component space, reducing each sample into one point. Thereby, we can separate samples by expression variation, and identify potential sample outliers. The PCA plot is a way to look at how samples are clustering.
```{r plot_pca}
label <- !identical_samples
plotPCA(
  object = bcb,
  normalized = "vst",
  label = label
)
```

## Covariates correlation with PCs

When multiple factors may influence the results of a given experiment, it is useful to assess which of them is responsible for the most variance as determined by PCA. We adapted the method described by Daily et al. where they integrated a method to correlate covariates with principal components values to determine the importance of each factor.

Here we are showing the correlational analysis of the rlog transformed count data's principal components with the metadata covariates of interest. Significant correlations (FDR < 0.1) are shaded from blue (anti-correlated) to orange (correlated), with non-significant correlations shaded in gray.

```{r plot_pca_covariates}
plotPCACovariates(bcb)
```

```{r plot_correlation_heatmap, eval=!identical_samples}
mdHeader("Hierarchical clustering", level = 2)
asis_output("Inter-correlation analysis (ICA) is another way to look at how well samples cluster by plotting the correlation between the expression profiles of the samples.")
plotCorrelationHeatmap(bcb, normalized = "vst")
```

# Save additional data

## Quality metrics

```{r metrics}
metrics <- metrics(bcb)
write.csv(
  x = metrics,
  file = file.path(params$results_dir, "metrics.csv")
)
```

## Gene annotations

```{r gene_annotations}
gene_annotations <- rowRanges(bcb) %>% as.data.frame()
write.csv(
  x = gene_annotations,
  file = file.path(params$results_dir, "gene_annotations.csv")
)
```

```{r footer, child="_footer.Rmd"}

```
